Loan Data from Propser by Tan Kwok Wee
========================================================
```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using
# in your analysis in this code chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk.
# This prevents the code from displaying in the knitted HTML output.
# You should set echo=FALSE for all code chunks in your file.

library(ggplot2)
library(GGally)
library(maps)
library(mapproj)
library(raster)
library(dplyr)
library(reshape)
library(plyr)
library(chron)
library(scales)
library(Hmisc)
library(lattice)

```

# Load the Data
```{r echo=FALSE}

ld <- read.csv("C:\\Users\\212369277\\Downloads\\prosperLoanData.csv")
ld$LoanOriginationDate <- as.Date(ld$LoanOriginationDate)
ld$LoanOriginationYear <- years(ld$LoanOriginationDate)
ld$LoanOriginationMonth <- as.factor(months(ld$LoanOriginationDate))
ld$LoanOriginationMonth <- factor(ld$LoanOriginationMonth,levels=c("January","February","March","April","May","June","July","August","September","October","November","December"))
ld$BorrowerState  <- as.character(ld$BorrowerState )
states_map <- map_data("state")
```

# Summary of data
The dataset consists of 113,937 rows. For each row there are 84 varibles.
```{r}

length(ld)
length(ld$ListingKey)

```
The below details the names, type and contents of the 84 variables
```{r}
str(ld)
```
the below provides a summary of the 84 varibles

```{r}
summary(ld)
```
Next lets investigate, the levels defined for EmployementStatus, CreditGrade,Occupation, loanstatus and range for loanOriginationDate to understand the levels for each variable

```{r}
levels(ld$EmploymentStatus)
levels(ld$CreditGrade)
levels(ld$Occupation)
levels(ld$LoanStatus)
range(ld$LoanOriginationDate)

```


# Univariate Plots Section

I start by checking the loan status of the borrowers in the dataset.

## Loan Status
```{r echo=FALSE}

ggplot(data=ld, aes(x=LoanStatus)) + geom_bar()  + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + stat_bin(geom="text", aes(label=..count.., vjust=-1)) + stat_bin(geom="text", aes(label=..count.., vjust=-1)) 

```

The dataset captures loans taken from 15-Nov-2005 to 12-Mar-2014. 
In the dataset 56,576 borrowers are servicing the loans and 38,074 completed their loans.

From the plot we can see that 5,018 - defaulted, 11,992 - Charged off and 2,064 (16+806+265+363+313+304)loans Past due date. In total, 19,074 loans are either defaulted, charged-off or past date due, constiuting 16.7% of the total loans in the dataset. 

Next, let's review the loan amount requested.

## Loan Original Amount
```{r echo=FALSE}
ggplot(data=ld, aes(x=LoanOriginalAmount)) + geom_density()

ggplot(data=ld, aes(x=LoanOriginalAmount)) + geom_density() + scale_x_continuous(limits=c(1000,20000),breaks=seq(0,20000,1000)) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 
```

By changing the limiting the x-axis scale, we can see that most loans taken are about $4000, $10000 and $15000. This shows that people like to borrow in round amounts 

next, let us understand the borrower rate.

## Borrower Rate
```{r echo=FALSE}

ggplot(data=ld,aes(x=BorrowerRate*100)) + geom_bar()

ggplot(data=ld,aes(x=BorrowerRate*100)) + scale_x_continuous(limits=c(0,38),breaks=seq(0,38,1)) + geom_bar() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + stat_bin(geom="text", aes(label=..count.., vjust=-1)) 


summary(ld$BorrowerRate)

```

by changing the breaks of the x-axis scale, we can determine that most of the borrowers borrowed at rate of 14 percent. the mean borrowing rate is 18.4 percent and  the max. borrowing rate is 49.75 percent. I guess that the rate at which the borrowers get their loan might depend on their employement status, credit profile and some other factors.

let's check the employment status of the borrowers

## Employment Status
```{r echo=FALSE}

ggplot(data=ld, aes(x=EmploymentStatus)) + geom_bar() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + stat_bin(geom="text", aes(label=..count.., vjust=-1)) + ylim(0,80000) + stat_bin(geom="text", aes(label=..count.., vjust=-1)) 
```

2,255 borrowers have no employment status, 5,347 with employment status as "Not available" and 835 not employed. It is interesting that 8,437 people having "Not available" or "Not Employed" as employement status obtaining loans.

Lets check income range of the borrowers in the dataset. 

## Income Range - General
```{r echo=FALSE}
ggplot(data=ld, aes(x=IncomeRange)) + geom_bar() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + stat_bin(geom="text", aes(label=..count.., vjust=-1)) 
```
In general,most of the borrower's income ranges from 25,000 to 75,000. It was suprisingly to see 621 borrowers having $0 income obtaining loans.

next let's review the monthly stated income

## Stated Monthly Income
```{r echo=FALSE}
ggplot(data=ld, aes(x=StatedMonthlyIncome)) + geom_bar() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

summary(ld$StatedMonthlyIncome)

ggplot(data=ld, aes(x=StatedMonthlyIncome)) + geom_bar() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+ scale_x_continuous(limits=c(0,10000),breaks=seq(0,10000,500)) + stat_bin(geom="text", aes(label=..count.., vjust=-1)) 

```

by ignoring the outliers and limiting the x-axis to under 10000. we can now better see the distribution of the monthly income range. Most of the borrowers are having 4000 to 4250 as their monthly income.

next lets check the creditgrade of the borrowers profile.

## Credit Grade
```{r echo=FALSE}

ld$CreditGrade <- factor(ld$CreditGrade, levels=c("AA","A","B","C","D","E","HR","NC",""))


ggplot(data=ld, aes(x=CreditGrade)) + geom_bar() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + stat_bin(geom="text", aes(label=..count.., vjust=-1)) 

```

We can see that majority of the borrowers (84,934) or approximately 75% of the dataset are without a credit Grade. It has noted that the datafield is populated pre-2009 and only be available for those listing. Most of the borrowers are having "C" as creditGrade in the dataset.

so Lets check the ProsperRating..Alpha., which is the field used after jul-2009 to track borrower's credit rating

## Prosper Loan Rating Alpha
```{r echo=FALSE}

ld$ProsperRating..Alpha. <- factor(ld$ProsperRating..Alpha., levels=c("AA","A","B","C","D","E","HR","NC",""))

ggplot(data=ld, aes(x=ProsperRating..Alpha.)) + geom_bar() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + stat_bin(geom="text", aes(label=..count.., vjust=-1)) 
 
ld.withoutCredit <- subset(ld, ld$ProsperRating..Alpha. =="" & ld$CreditGrade =="", select=c("LoanOriginationDate","ProsperRating..Alpha.","CreditGrade") )

count(ld.withoutCredit,"LoanOriginationDate")
```

from the plot we can see that The median prosper rating alpha is C.
We can see that 29,084 loans are without prosper rating alpha. 
In total 84,853 borrowers are assigned with prosper rating alpha.

we also can see that there are 99 borrowers are without ProsperRating..Alpha. or Credit Grade, and the occurence is consistently spread over 2009 to 2010.

let us check if the borrowers are over-leveraged by checking debit-to-income ratio 

## Debt to Income Ratio
```{r echo=FALSE}

ggplot(data=ld, aes(x=DebtToIncomeRatio)) + geom_bar() + stat_bin(geom="text", aes(label=..count.., vjust=-1)) 

ggplot(data=ld, aes(x=DebtToIncomeRatio)) + geom_bar(bindwidth=1) + scale_x_continuous(limits=c(0,2.5),breaks=seq(0,2.5,0.1)) + scale_y_log10() + stat_bin(geom="text", aes(label=..count.., vjust=-1)) 

summary(ld$DebtToIncomeRatio)
```

from the plot, we can see that most of the borrowers are having debt to income ratio below 1.0


there are some outlier data, showing that borrowers with debt to income ratio of 10.01.

I had limited the maximium Debt to Income ratio to be 2.5 and re-plotted the data. We can see a left skew distribution and most of the borrowers have debt to income ratio of 0.2.

## Term
```{r echo=FALSE}

ggplot(data=ld, aes(x=Term)) + geom_bar()

ggplot(data=ld, aes(x=Term)) + geom_bar() + scale_x_continuous(limits=c(10,65),breaks=seq(10,65,5)) 
```

we can see that most of the loans are taken with term of 36 months.
lets check when the loans happened

## Loan Origination Year
```{r echo=FALSE}

ggplot(data=ld, aes(x=LoanOriginationYear)) + geom_bar() 

summary(ld$LoanOriginationYear)
```

We can see that there is significant drop in loans taken after 2008.
After 2009, more people are taking loans. 

let's check if there is people take more loans in particular months by creating a plot of Loan Oringation Date by formatting the date as month

# Loan Origination Month
```{r echo=FALSE}
ggplot(data=ld, aes(x=LoanOriginationMonth)) + geom_bar() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + stat_bin(geom="text", aes(label=..count.., vjust=-1)) 

summary(ld$LoanOriginationMonth)
```

We can see that March, April, May has the lowest number of loan taken.
Borrowers then to take more loans in Janurary and October.
Let's check if we modify the loan origination date as month-year date format

## Loan Origination Date
```{r echo=FALSE}

ggplot(data=ld, aes(x=LoanOriginationDate)) + geom_bar() + scale_x_date(breaks = date_breaks("months"),                                                                  labels = date_format("%b-%y")) + theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1))

summary(as.factor(format(ld$LoanOriginationDate,format="%b-%y")))

```

plotting the loan oringination date using month-year format, we can see an increase in loan happened after May 2013. I wonder if there was any specifc event that cause the rapid increase of loans taken?

let check the states where the borrower's live and their occupation next.

## Borrower State
```{r echo=FALSE}

ggplot(data=ld, aes(x=BorrowerState)) + geom_bar() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

we can see that most of the borrowers live in CA, NY, TX, FL and IL

## Occupation
```{r echo=FALSE}

ggplot(data=ld, aes(x=Occupation)) + geom_bar() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

Top 5 borrowers Occupation are professional (13628), computer programmer (4478) and executive (4311), Teacher (3759) and administrative assistant (3588)

let us check what are the top reasons for taking loans

## Listing Category
```{r echo=FALSE}
ggplot(data=ld, aes(x=ListingCategory..numeric.)) + geom_bar() + scale_x_continuous(breaks=seq(0,20,1), labels=c("Not Available", "Debt Consolidation", "Home Improvement","Business","Personal Loan","Student Use","Auto","Other","Baby&Adoption","Boat","Cosmetic Procedure","Engagement Ring","Green Loans","Household Expenses","Large Purchases","Medical/Dental","Motorcycle","RV","Taxes","Vaction","Wedding Loans")) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + stat_bin(geom="text", aes(label=..count.., vjust=-1)) 
```

we can see that majority of the loans are for Debt Consolidation (58,308), Others(10,494) and Home Improvement (7,433).

## Total Proper Loans
```{r echo=FALSE}

ggplot(data=ld, aes(x=TotalProsperLoans)) + geom_bar(bindwidth=1)+ stat_bin(geom="text", aes(label=..count.., vjust=-1)) + stat_bin(geom="text", aes(label=..count.., vjust=-1)) 
```

we can understand from the plot that most of the borrowers have only 1 prosper loan.

# Univariate Analysis

### What is the structure of your dataset?
There are 113,937 listings in the dataset with 83 features. I had chosen to analyze the below stated features 

LoanOriginalAmount
, Credit Grade or ProsperRating..Alpha.
, stated monthly income
, Debt to Income Ratio
, Total Prosper Loans
, Current Delinquencies
, Delinquencies past 7 years
, Borrower Rate
, Income Range
, LoanOriginationDate
, Occupation
, Employment Status

below describes the ranking of levels of the features explored

(Worst to Best)

Credit Grade : NC HR E D C B A AA

Income Range : Not employed $0 $1-$24,999 $25,000-$49,999 $50,000 - $74,999 $75,000 - $99,999 $100,000+ Not displayed

ProsperScore Alpha : NC HR E D C B A AA

### What is/are the main feature(s) of interest in your dataset?

The main features of the data set are CreditGrade/Propser Score Alpha, Borrower Rate, Debt to income ratio, Monthly Stated Income and Total prosper loan. 

I will want to explore what features will impact the borrower rate.

### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?

I believe related features like Current Delinquencies
, Delinquencies past 7 years can help support my investigation

### Did you create any new variables from existing variables in the dataset?

I created variable LoanOriginationYear by formatting the LoanOriginationDate to display year.

I create varaible LoanOriginationMOnth by formatting the LoanOriginationDate to display month.

### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?

I found that feature Loan Orgination date has unusual distribution.

when ploting the Loan Origination month histogram, i can see a slight "concave" distribution. Therefore, i converted the month into month-year date format and re-plotted the histogram. We can now see that the month-year date format depicts distribution skewed towards the right. We can determine that there was an increased amount of loans are taken May 2013

# Bivariate Plots Section

to get an overview of the relationship between identified featuresets, i use ggpairs to generate a summary of plots of features on a small sample size of 10,000 records.

## Correlation Analysis

```{r echo=FALSE}

set.seed(20150508)
ld.sample <- ld[sample(1:length(ld$LoanOriginalAmount),10000),]
ld.sample[is.na(ld.sample)] <- 0

ld.sample <- subset(ld.sample, select =c(DebtToIncomeRatio,TotalProsperLoans,BankcardUtilization,StatedMonthlyIncome,BorrowerRate,LenderYield,LoanOriginalAmount,AmountDelinquent,DelinquenciesLast7Years,RevolvingCreditBalance,OnTimeProsperPayments,CurrentDelinquencies,CurrentCreditLines,ProsperPrincipalBorrowed,PublicRecordsLast10Years,EstimatedLoss))

ggpairs(ld.sample)

cor(ld.sample)

```

From the plot, we can see a strong correlation between borrower rate and lender yield (0.99), Proper Principal Borrowed with Total prosper loans (0.81) and  on time propser loan payments with prosper principal borrowered (0.69).

let's focus on borrowers income range for borrowers with emmployment status as "not employed" or "no employment status".

## Income Range - employment status = Not Employed or Not available
```{r echo=FALSE}
levels(ld$EmploymentStatus)

ggplot(data=subset(ld,ld$EmploymentStatus %in% c("Not employed","Not available","")), aes(x=IncomeRange)) + geom_bar() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + stat_bin(geom="text", aes(label=..count.., vjust=-1)) + facet_wrap(~EmploymentStatus)

```

if we subset the data for those borrowers without employment status or not available and or employed, we uncover some interesting details.

there are 11 people in this group declared they have $0 as well as having employment status as not employed obtaining loans. 

there are 806 people in this group declared their income range as "not employed"" for income range as well as not employed obtaining loans. 

Most of the people with income range status as "Not displayed" come from the group of people declaring their employment status as (no employment status/not available or not employed).

I wanted to know if term changes with the loan amount and decided to do a violin plot to understand the distribution.

## Term vs Loan Original Amount
```{r echo=FALSE}
ggplot(data=ld, aes(x=Term,y=LoanOriginalAmount)) + geom_violin()

count(subset(ld,select=c("Term")), vars=c("Term"))

head(arrange(count(subset(ld,select=c("Term","LoanOriginalAmount")), vars=c("Term","LoanOriginalAmount")), desc(freq)))

```

we can see that for loan amounts under 10,000, the term varies widely.
for loans above 30,000, we can see the term varies little.

by constructing a frequency table by term, we can see that most people opt for 36 months loans, followed by 60 months loan.

by constructing a frequency table by term and loan amount, we can see that most people opt for 36 months loans for loan of 4,000 and followed by 36 months loan for loan of 10,000.

## Loan Origination Year Vs Loan Original Amount
```{r echo=FALSE}

ggplot(data=ld, aes(x=LoanOriginationYear,y=LoanOriginalAmount)) + geom_boxplot()

head(arrange(count(subset(ld,select=c("LoanOriginationYear","LoanOriginalAmount")), vars=c("LoanOriginationYear","LoanOriginalAmount")), desc(freq)))

head(arrange(count(subset(ld,select=c("LoanOriginationYear","LoanOriginalAmount"), LoanOriginationYear < 2011), vars=c("LoanOriginationYear","LoanOriginalAmount")), desc(freq)))

```

We can see that range of loans amount increases after 2011. 
Most loan amounts requested are 15,000, 4,000 and 10,000 after 2011.
Most loan amounts requested are 5000 and 1000 before 2011.

i am curious to understand which state does the bulk of the borrowwers reside in.

## Borrower State
```{r echo=FALSE}

ld$state <- tolower(state.name[match(ld$BorrowerState,state.abb)])
ld.state <-subset(ld, select=c("state"))
ld.state <- count(ld.state,"state")

ld.state[with(ld.state,order(- freq, state)),]

ld.state <- merge(states_map, ld.state, by.x="region", by.y="state")

ld.state <- arrange(ld.state,group,order)

snames <- data.frame(region=tolower(state.name), long=state.center$x, lat=state.center$y)

snames <- merge(snames, subset(ld.state,select=c("region","freq","group")), by="region")

snames <- distinct(snames)

ggplot(ld.state, aes(x=long,y=lat, group=group, fill=freq)) + geom_polygon(color="black") + scale_fill_gradient2(low="#559999",mid="blue", high="red") + coord_map("polyconic") + geom_text(data=snames, aes(long, lat, label=freq))

ggplot(ld.state, aes(x=long,y=lat, group=group, fill=freq)) + geom_polygon(color="black") + scale_fill_gradient2(low="#559999",mid="blue", high="red") + coord_map("polyconic") + geom_text(data=snames, aes(long, lat, label=region))

```

It is clear that the bulk of borrowers live in California, Texas, New York Florida and Illinois.

States like Montana, Wyomin, North Dakota, South Dakota and Maine has the least amnount of borrowers.

I am curious to see the distribution of credit grade and loan amounts of borrowers living in California, Texas, New York Florida and Illinois. 

## Borrower State ( CA, TX, NY, IL, FL ) and Percentage of CreditGrade
```{r echo=FALSE}
ld.state.creditgrade <- subset(ld, BorrowerState %in% c("CA","TX","NY","IL","FL") ,select=c("CreditGrade", "BorrowerState","LoanOriginalAmount") )

ld.state.creditgrade <- subset(ld.state.creditgrade,CreditGrade %in% c("AA","A","B","C","D","E","HR","NC"))

ld.state.creditgrade$CreditGrade <- factor(ld.state.creditgrade$CreditGrade, levels=c("AA","A","B","C","D","E","HR","NC"))

ld.state.creditgrade$BorrowerState <- factor(ld.state.creditgrade$BorrowerState)

ld.state.creditgrade <- ddply(ld.state.creditgrade, "BorrowerState", transform,row_state_total=length(BorrowerState),row_state_amt_total=sum(LoanOriginalAmount))

ld.state.creditgrade <- ddply(ld.state.creditgrade, c("BorrowerState","CreditGrade"), transform,row_state_CreditGrade_total=length(CreditGrade),row_state_CreditGrade_amt_total=sum(LoanOriginalAmount))

ld.state.creditgrade <- distinct(ld.state.creditgrade[,c("CreditGrade","BorrowerState","row_state_total","row_state_amt_total","row_state_CreditGrade_total","row_state_CreditGrade_amt_total")])

ld.state.creditgrade$percent <- ld.state.creditgrade$row_state_CreditGrade_total/ld.state.creditgrade$row_state_total

ld.state.creditgrade$percent_amt <- ld.state.creditgrade$row_state_CreditGrade_amt_total/ld.state.creditgrade$row_state_amt_total

ggplot(data=ld.state.creditgrade, aes(x=BorrowerState,y=percent,fill=CreditGrade)) + geom_bar(stat="identity") 
```

New York has the highest amount of borrowers with credit grade of AA and A, and i suspect that people there has better paying jobs then rest of the states in the histogram.

## Credit Grade vs Loan Original Amount, Proper Rating Alpha vs Loan Amounts
```{r echo=FALSE}

ggplot(data=subset(ld,CreditGrade !=""), aes(x=CreditGrade,y=LoanOriginalAmount)) + geom_boxplot()

ggplot(data=subset(ld,ProsperRating..Alpha.!=""), aes(x=ProsperRating..Alpha.,y=LoanOriginalAmount)) + geom_boxplot()
```

In general, we can confirm that people with higher credit grade will be able to obtain higher loan amounts.

## Credit Grade/ProsperRating..Alpha. vs BorrowerRate 
```{r echo=FALSE}

ggplot(data=subset(ld,CreditGrade !=""), aes(x=CreditGrade,y=BorrowerRate)) + geom_boxplot()

ggplot(data=subset(ld,ProsperRating..Alpha.!=""), aes(x=ProsperRating..Alpha.,y=BorrowerRate)) + geom_boxplot()

```

by plotting GreditGrade versus Borrower Rate we can see that borrowers with lower credit grade paid more interest as compared with borrowers with higher credit grade. for borrowers with lower creditgrade the Borrower Rates varies wider as compared with those having a better creditgrade

## Employment Status vs BorrowerRate 
```{r echo=FALSE}

ggplot(data=ld, aes(x=EmploymentStatus,y=BorrowerRate)) + geom_boxplot()

```

We can see that borrowers that are "Not Employed" and "Other" have a higher borrower rate ( greater than 20%) as compared with other employment status.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?

we can see as loan amounts get larger the term of loan stablizes around 36 months.

we can see that the range of loan amounts increases after 2011

If we see the boxplots of credit grade verus Loan Origination Amount, we can see that highest credit Grade obtained more loan amounts at lower borrower rate. 

There are certain outliers obtaining loan amounts of 20,000 or more despite having a bad credit rating.

We can see that people that are employed requested for bigger loan amounts as compared with other employment status.

There is a linear relationship between borrower rate and lender yield, increase in borrower rate corresponds with an increase with lender yield.

we can also see that borrowers with the Best credit grade enjoy lowest borrower rate.

### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?

In the plot of BorrowerState vs CreditGrade, we can observe that people in NY has better credit grade than the rest of the states ( CA, FL, IL, TX)

### What was the strongest relationship you found?

the strongest relationships i found is between borrower yield and lender rate
they have the correlation of 99.9%

# Multivariate Plots Section

Let us check the distirbution of loan amounts by Credit Grade and borrower State for (CA, TX, NY, IL, FL)

## Borrower State ( CA, TX, NY, IL, FL ) and CreditGrade and Percentage of Loan Amount 
```{r echo=FALSE}
ggplot(data=ld.state.creditgrade, aes(x=BorrowerState,y=percent_amt,fill=CreditGrade)) + geom_bar(stat="identity") 

```

We can see that in terms of credit grade and loan amount percentage Florida and New York have almost the same distribution.

Next let us check the borrower Rate vs Credit Grade by IncomeRange

## Borrower Rate/Prosper Rating Alpha vs Credit Grade/ProsperRating..Alpha. by Income Range
```{r echo=FALSE}
ld$EmploymentStatus <- factor(ld$EmploymentStatus, levels=c("Employed","Full-time","Part-time","Self-employed","Retired","Other","Not employed","Not available",""))

ggplot(data=subset(ld, ld$CreditGrade != "") , aes(x=BorrowerRate, fill=CreditGrade)) + geom_bar() + facet_wrap(~IncomeRange, ncol=2)

ggplot(data=subset(ld, ld$ProsperRating..Alpha. != "") , aes(x=BorrowerRate, fill=ProsperRating..Alpha.)) + geom_bar() + facet_wrap(~IncomeRange, ncol=2)

```

we can see the similar patterns of distribution of borrwing rates across all income range which varies by the CreditGrade/ProsperRating..Alpha.. The borrowers having lower creditgrade or prosperrating..alpha. (D,E,HR) are having interest rates greater than 0.2

## Employment Status vs Borrower Rate by Credit Grade/ProsperRating..Alpha.
```{r echo=FALSE}
ggplot(data=ld, aes(x=EmploymentStatus,y=BorrowerRate*100, fill=CreditGrade)) + geom_boxplot() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

ggplot(data=ld, aes(x=EmploymentStatus,y=BorrowerRate*100, fill=ProsperRating..Alpha.)) + geom_boxplot() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

we can see that even if group modify the plot to cateogrize by employment status, the highest credit Grade/ProposerRating..Alpa. (AA) has the lowest borrowing rate accross all Employment status. 

Next, let's understand loss by employment status grouped by CreditGrade/Proper Rating Alpha

## Employment Status vs Loss by Credit Grade

```{r echo=FALSE}
ggplot(data=ld, aes(x=EmploymentStatus,y=EstimatedLoss, fill=ProsperRating..Alpha.)) + geom_boxplot() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

We can see that the estimated loss for different employement status by credit grade have almost the same variation and borrowers with lower credit grade will have a higher estimated loss percentage. lets check if higher loan amounts will mean higher estimated loss?

## Loan Amount vs Estimated Loss by Proper Rating .. Alpha
```{r echo=FALSE}

coef(lm(EstimatedLoss*100 ~ LoanOriginalAmount, data = ld))

ggplot(data=ld, aes(x=LoanOriginalAmount,y=EstimatedLoss*100, color=ProsperRating..Alpha.)) + geom_point() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + geom_abline(intercept = 10.9346917298, slope = -0.0003197143)

```

there is a negative correlation between esitmated loss vs loan original amount.
let's check estimated loss with debt to income ratio

## Debt to Income Ratio vs Estimated Loss by Proper Rating .. Alpha
```{r echo=FALSE}

coef(lm(EstimatedLoss*100 ~ DebtToIncomeRatio, data = ld))

ggplot(data=ld, aes(x=DebtToIncomeRatio,y=EstimatedLoss*100, color=ProsperRating..Alpha.)) + geom_point() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + geom_abline(intercept = 7.388626, slope = 1.82)

```

We can see a positive correlation between Estimated loss vs Debt to Income Ratio. We also can see borrowers with poor Prosper Rating.. Alpha (D, E, HR) has higher estimated loss. There are more occurences when debt to income ratio is greater than 2.0.

## Linear Model - Borrower Rates
```{r echo=FALSE}
ld.lm_br <- lm(formula = BorrowerRate ~  ProsperScore + TotalProsperLoans + ProsperRating..Alpha. , data=ld)

summary(ld.lm_br)
```

We can see that prosper score, total prosper loans and propserrating.. alpha are the independent variables that are able to explain 91.59% of the variation in borrower's rate

## Linear Model - Estimated Loss
```{r echo=FALSE}
ld.lm_el <- lm(formula = EstimatedLoss ~  ProsperScore + TotalProsperLoans + ProsperRating..Alpha. , data=ld)

summary(ld.lm_el)

```

Similarly, prosper score, total prosper loans and propserrating.. alpha are the independent variables that are able to explain 92.38% of the variation in estimated loss.

## Linear Model - Lender Yield
```{r echo=FALSE}
ld.lm_ly <- lm(formula = LenderYield ~  ProsperScore + TotalProsperLoans + ProsperRating..Alpha. , data=ld)

summary(ld.lm_ly)

```

Similarly, prosper score, total prosper loans and propserrating.. alpha are the independent variables that are able to explain 91.59% of the variation in lender yield.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature(s) of interest?

We can see a positive correlation between Estimated loss vs Debt to Income Ratio. We also can see borrowers with poor Prosper Rating.. Alpha (D, E, HR) has higher estimated loss. There are more occurences when debt to income ratio is greater than 2.0.

### Were there any interesting or surprising interactions between features?

we can observe that there are people without stated monthly income with loans greater than 20,000

There is a negative correlation between esitmated loss vs loan original amount.
My initial impression that greater loan amounts will affect estimated loss percentage.

### OPTIONAL: Did you create any models with your dataset? Discuss the strengths and limitations of your model.

yes, i had created 3 linear models on Borrower Rates, Estimated Loss and Lender Yield.

I had used multiple linear regression model to predict dependant variable ( BorrowerRate, EstimatedLoss) and using Adjusted R-Square to evaluate the model.

The model is simple to use and understand, but linear regression is unable to predict outliers and unable to establish causation.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE}
ggplot(data=ld, aes(x=ListingCategory..numeric.,fill=ProsperRating..Alpha.)) + geom_bar() + scale_x_continuous(breaks=seq(0,20,1), labels=c("Not Available", "Debt Consolidation", "Home Improvement","Business","Personal Loan","Student Use","Auto","Other","Baby&Adoption","Boat","Cosmetic Procedure","Engagement Ring","Green Loans","Household Expenses","Large Purchases","Medical/Dental","Motorcycle","RV","Taxes","Vaction","Wedding Loans")) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + facet_wrap(~EmploymentStatus, ncol=1,scales="free_y")
```

### Description One

By plotting Listing Category by ProsperRating..Alpha and facet by Employment Status. We can see majority of the Listing Category is "NA" for all different employment status. We can also see that those without Prosper Rating..Alpha constitues the majority in the Listing Category is "NA".

Debt Consolidation is primary reason after "NA" for borrowing accross most employment status. 

From the plot, we can observe the following
* borrowing for "Business"" purpose has more occurences in "self employed" group then the rest 
* borrowing for "Student Use" more common in "Part-time" group 
* borrowing for "Home improvement" is more common in "retired" group

### Plot Two
```{r echo=FALSE}

ld$state <- tolower(state.name[match(ld$BorrowerState,state.abb)])
ld.state <- aggregate(StatedMonthlyIncome ~ state, data=ld, mean)


ld.state <- merge(states_map, ld.state, by.x="region", by.y="state")

ld.state <- arrange(ld.state,group,order)

snames <- data.frame(region=tolower(state.name), long=state.center$x, lat=state.center$y)

snames <- merge(snames, subset(ld.state,select=c("region","StatedMonthlyIncome","group")), by="region")

snames <- distinct(snames)

ggplot(ld.state, aes(x=long,y=lat, group=group, fill=StatedMonthlyIncome)) + geom_polygon(color="black") + scale_fill_gradient2(low="#559999",mid="blue", high="red") + coord_map("polyconic") + geom_text(data=snames, aes(long, lat, label=region))

arrange(distinct(subset(snames,select=c("region","StatedMonthlyIncome"))), desc(StatedMonthlyIncome), region)

```

### Description Two

We can see that Connecticut, New Jersey, Maryland, Virginia and New York are the top 5 states with highest mean stated Monthly income. When we compare with the number of loans taken (where the bulk of borrowers live in California, Texas, New York, Florida and Illinois) we do not see there is a 100% correlation between top 5 loans borrowed by state vs top 5 mean statedMonthlyIncome by state. If we check the state population estimate as of 1st July 2014 from wikipedia, we can see that the top 5 states with the biggest population corresponds with the top 5 states with number of loans. 

### Plot Three
```{r echo=FALSE}

ggplot(data=subset(ld,ProsperRating..Alpha.!=""&DebtToIncomeRatio<1.0), aes(x=DebtToIncomeRatio,y=LenderYield*100, color=ProsperRating..Alpha.)) + geom_point() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 

```

### Description Three

Based on the plot, we can conclude that lower proper rating is associated with higher yield to lender despite the borrower having a low Debt To Income Ratio <= 1.0. We can see that borrower the same proper rating clustered aroung the same ranges of yield.

```{r}
range(arrange(distinct(subset(ld,ProsperRating..Alpha.=="AA"&DebtToIncomeRatio<1.0,select=c("LenderYield"))),desc(LenderYield)))
```

We can see that lender yield ranges from 3% to 20% for Borrowers with propser rating alpha of "AA"" and debt to income ratio below 1.0.

------

# Reflection

After doing this project that i realize that R is a powerful visualization and statistical computation tool. It is very easy to compute statistics and graphs with the help of libraries like plyr, dylr and ggplot.

Based on the models created we can see that 3 variables prosper score, total prosper loans and propserrating.. alpha has correlation of over 90% with BorrowerRate, Estimated Loss and Lender yield. I feel that, it will be interesting to build a model to determine prosper score to enrich the analysis.

Personally, I feel that Prosper Loans is a viable alternative investment. Based on my analysis of the borrowers with Prosper Rating .. Alpha of "AA" and Debt to income Ratio below 1.0 the lender yield is between 3% to 20% which I feel is an attractive return.

After this project, i feel that i equiped with the necessary knowledge to explore other datasets.

# Reference

* [How To Find Relationship Between Variables, Multiple Regression] (http://www.statsoft.com/textbook/multiple-regression)
* [Prosper - company Overview] (https://www.prosper.com/about)
* [GGplot from yhat] (http://ggplot.yhathq.com/)
* [R Studio - rmarkdown reference](http://www.rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf)
* [R Graphics Cookbook] (http://www.cookbook-r.com/Graphs/index.html)
* [Wikipedia - List of U.S. states and territories by population](http://en.wikipedia.org/wiki/List_of_U.S._states_and_territories_by_population)
